Comparison of two object-oriented languages: C++ and Ruby

### 2012-01-15 paper.tex
Final version of the paper. Needs finalfinal check :p

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{color}
\definecolor{codebg}{rgb}{0.9,0.9,0.9}
\usepackage{listings}
\lstset{language=C++,
backgroundcolor=\color{codebg},
numbers=left,
xleftmargin=9pt,
framexleftmargin=9pt,
basicstyle=\scriptsize,
numberstyle=\scriptsize\color{gray},
breaklines=true,
numbersep=1pt,
tabsize=2}
\newenvironment{quoteje} 
               { \noindent {\bfseries  "}}     
               {{\bfseries  "}\vspace{\parskip}}
\renewcommand\refname{References}
\author{Jeroen Heymans}
\title{Comparison of two object-oriented languages: C++ and Ruby}
\begin{document}
\maketitle

@@

\columnbreak

@@

\subsubsection{Ruby}
When designing Ruby, the creator of Ruby, Ykihiro Matsumoto, looked at other languages to find an ideal syntax. He wanted a dynamic typed scripting language that was more powerful than Perl and more object-oriented than Python. In Ruby, everything is an object. It was influenced by Smalltalk, every type was given method instance variables. All Ruby code was tested with Ruby version 1.9.2p0 (2010-08-18 revision 29036).

\subsection{What follows next}
We will compare several important features that belong to object-oriented languages like inheritance, polymorphism, access control and many more.

\pagebreak
\section{Comparisons}

@@

Due to the dynamic typed nature of Ruby, there is no need for constructions like templates in C++ to enforce one function or method to accept multiple types.

\subsection{Classes}
In object-oriented programming, a class is a blueprint that is used to construct objects which we call object instances. We compare the several types of classes that can be made in Ruby and C++.

\subsubsection{Abstract classes}
An abstract class is a class that can not be instantiated for it is either labelled as abstract or it specifies abstract methods. The general idea of abstract classes is that the class must be extended. In C++, an abstract class is a class having at least one pure virtual function. For example, this is an abstract class in C++:

\begin{lstlisting}
class A {
	public:
		virtual void f() = 0;
};
\end{lstlisting}

If we are trying to compile C++ code where instantiate this abstract class, for example like this:

\begin{lstlisting}
int main() {
   A a = new A();
   return 0;
}
\end{lstlisting}

We will receive an error message from the compiler that notifies us that this operation is not allowed:

@@

Abstract classes in Ruby are a different story. Unlike some object-oriented languages, Ruby does not support a keyword like \textit{abstract} or a way like C++ to define an abstract class. Actually, Ruby does not provide the necessary tools to create abstract classes. However, the programmer himself can force a class to behave like an abstract class when he makes \textit{:new} (the constructor) private. The extending class must then make \textit{:new} public:

\begin{lstlisting}[language=Ruby]
class A
	private_class_method :new
	def initialize(txt = "Hello!")
		puts txt
	end
end

class B < A
	public_class_method :new
end
\end{lstlisting}

As long as there is no class that extends A, the methods in A will never be accesible as it is not possible to construct an instance of A.

\subsubsection{Final classes}
A final class is a class that can not be extended. This could be done for reasons of security and efficiency. Final classes are not supported in C++ nor Ruby however there exist mechanisms to prevent programmers from extending other classes. An example of such a mechanism in C++:

\begin{lstlisting}
class Final;

class MakeFinal {
	private:
    	MakeFinal() { }
    friend class Final;
};

class Final : virtual MakeFinal {
	public:
		Final() { }
};

class Derived : public Final { };
\end{lstlisting}

In this example, we made the class named Final final. We did this by inheriting class Final from the class MakeFinal. In class MakeFinal, we have a private constructor which can only be called by the class Final as the class Final is marked as \textit{friend}. By inheriting MakeFinal \textit{virtual} in the class Final, we make sure that the constructor \textit{MakeFinal()} must be called if we derive the class Final. As only Final is a friend, this will generate a compile error that notifies us that \textit{MakeFinal()} is private.

In Ruby, we can create a similar mechanism:

\begin{lstlisting}
class Class
   def final(klass)
      class << klass
         def inherited(subclass)
            raise RuntimeError, "Illegal attempt to subclass #{self} with #{subclass}"
         end
      end
   end
end

class Foo
   final(self)
end

class Bar < Foo
end
\end{lstlisting}

In this example, we add a method \textit{final} to the standardclass \textit{Class}. In this method \textit{final}, we define that a new method called \textit{inherited} should be added to the class given by the parameter \textit{klass}. \textit{inherited} is a method that will always be called whenever we try to inherit the class that contains this method. In our example, this will cause a \textit{RuntimeError}, thus disallowing the programmer to continue programming as long as we try to inherit our \textit{final} class.

But is a class really final in Ruby if we force a class to be final like we did in C++? Unfortunately no. This is due to the \textit{open classes} concept of Ruby that will always allow you to add methods to a class, whether it is marked as \textit{final} like in our example or not.\footnote{See the section "Partial classes" however for an example on how to prevent extending an open class}

@@

\begin{itemize}
\item Friendships are not corresponded: if class A is a friend of class B, class B is not automatically a friend of class A.
\item Friendships are not transitive: if class A is a friend of class B and class B is a friend of class C, then is class A not a friend of class C.
\item Friendships are not inherited: if there is a class Base with a friend, then the class Derived (that inherits from Base) does not have that same friend. The same goes up for all friends of Derived, they are not automatically friends of Base. The same goes in the other direction: if Base (or Derived) is a friend of a class, then is Derived (or Base) not automatically a friend of the same class.
\item The restricted members and methods from Base that are inherited by Derived are accessible by a friend class of Derived if Derived itself can access these members and methods. So if Derived inherits publicly\footnote{See the section "Derived classes" in "Access control" for more details on deriving classes.} from Base, Derived only has access to the protected and public members from Base, not the private members, so neither does a friend (the same goes for methods).

Friend classes are not possible in standard Ruby, however there exist some libraries\footnote{An example can be found on https://github.com/lsegal/friend} that allow a programmer to define friend classes in Ruby.
\end{itemize}

\subsubsection{Inner classes}
Inner classes are called nested classes in Ruby. They can simply be defined like you would expect:

\begin{lstlisting}[language=Ruby]
class A
	attr_accessor :a
	def initialize
		@a = "a"
	end
	def f
		puts @a
	end
	class B
		attr_accessor :b
		def initialize 
			@b = "b"
		end
		def g
			puts @a
		end
	end
end
\end{lstlisting}

We can then construct an object of the type \textit{A} and \textit{A::B}:

@@

With inner classes in C++, almost the same story exists. In order to access data from the inner class via the outer class or from the outer class via the inner class, an instance must be present to access this data. This can be achieved when constructing the inner or outer class:

\begin{lstlisting}
#include <iostream>

class A {
	private:
		int a;
	public:
		A(int newA): a(newA), b(this) { }
		class B {
			public:
				B(A* a): outer(a), test(1) { }
				int f() {
					return outer->a;
				}
				A* outer;
			public:
				int test;
		} b;
		int f() {
			return b.test;
		}
};

int main()
{
   A a = A(3);
   std::cout << a.b.f() << std::endl;
   std::cout << a.f() << std::endl;
   return 0;
}
\end{lstlisting}

@@

However, there are some rules concerning the access to and from the inner class in C++. The code above will compile and work. But if you look closely, the method \textit{f()} from the class B can access member \textit{a} from class A eventhough \textit{a} is defined as private. So in fact, the inner class gets the privilege to act as if it is part of the outer class.

We must notice though that there is a different story when you want to access the inner class. In the code above, A has a method \textit{f()} that access the member \textit{test} that is defined in B, via the pointer to B that was saved in A. This is because the member \textit{test} is defined as public. If we however define \textit{test} as protected or private, the compiler will notify us that this is not allowed.

@@

C++ does not have built-in support for metaclasses. However, it is possible to mimic the behaviour of meta-classes. This would mean however for example that you manage a metaclass yourself where the metaclass is a singleton class.

\subsubsection{Partial classes}
A partial class is a class whose definition may be splitted into multiple pieces. These pieces can be within one source file or across multiple files. Ruby supports the notion of partial classes. With this, it is possible to extend existing classes with new methods or members. The class must not be user-defined, it can be for example the built-in \textit{String} class:

\begin{lstlisting}[language=Ruby]
class String
	def doublePrint
		print self
		print self
	end
end
puts "test".doublePrint
\end{lstlisting}

This will give as a result: "\textit{testtest}".

Partial classes like in Ruby do not exist in C++. It is not possible to add new methods or members on-the-fly. The reason for this is that Ruby classes are considered "open" while C++ classes are "closed". When you define a C++ class in the typical header-file, there is no possibility to split this definition in several files, e.g. multiple header files for the same class are not allowed. Only the definition of the methods themselves can be split over mutiple source-files. It is for example possible to have a header-file "a.h" and multiple source-files like "method1-from-a.cpp" and "method2-from-a.cpp". Still, this does not create pure open classes like in Ruby where you can add methods at runtime, it is merely a way to make smaller files.

If you do not want that classes are further extended in Ruby, then the possibility of \textit{freezing} the class exists. By simple adding the keyword \textit{freeze} to the class, you declare the class as frozen (e.g. not modifiable).

@@

Ruby also provides these three keywords but there are some subtle differences. We will now explain one difference, the rest will be explained in the following section. The difference between Ruby and C++ can be found in the objects on which a private method is called. This C++ code compiles:

\begin{lstlisting}
class A {
	private:
		void foo() { return; }
	public:
		void test(A* a) {
			foo();
			this->foo();
			a->foo();
		}
};

int main()
{
   A* a = new A();
   A* b = new A();
   a->test(b);
   return 0;
}
\end{lstlisting}

As you can see, we have a class A in which the \textit{test()}-method accepts a parameter to an A object. All three calls to the method \textit{foo()} work: with an implicit receiver and an explicit receiver like \textit{this} or \textit{a}. If we try the Ruby equivalent of this code, we get into trouble:

\begin{lstlisting}
class A
	private
		def foo
		end
	public
		def test(obj)
			foo
			self.foo
			obj.foo
		end
end

a = A.new
b = A.new
a.test(b)
\end{lstlisting}

Both the lines \textit{self.foo} and \textit{obj.foo} generate the same error \textit{NoMethodError}. This is because we use an explicit receiver. The implicit receiver in line 7 allows the call to \textit{foo}. In Ruby, it depends on the type of receiver that you use while this is not important in C++.

\subsubsection{Derived classes}
When declaring a derived class in C++, we can provide an access specifier: \textit{public}, \textit{protected} or \textit{private}. For example if we have the base class A, we can create subclass B from A and provide one of the three access specifiers like this:

\begin{lstlisting}
class A { };
class B: public A { };
\end{lstlisting}

The purpose of this access specifier is to alter the access control of the members of the base class via the derived class. The access control is not altered on the original base class. An example:

\begin{lstlisting}
#include <iostream>

class A {
	public:
		int foo;
		A(): foo(0) {};
};

class B : public A { };

int main()
{
   B* b = new B();
   b->foo = 3;
   std::cout << "Value foo: " << b->foo << std::endl;
   return 0;
}
\end{lstlisting}

When compiling and excecuting this code, the program will output correctly:
\begin{lstlisting}
Value foo: 3
\end{lstlisting}

If we however change the line access keyword when defining the class B to for example \textit{protected} or \textit{private} (instead of the current \textit{public}), we get a compilation error eventhough everything in A was originally defined as \textit{public}:

\begin{lstlisting}
g++ a.cpp -o a
a.cpp: In function 'int main()':
a.cpp:5:7: error: 'int A::foo' is inaccessible
a.cpp:14:7: error: within this context
a.cpp:5:7: error: 'int A::foo' is inaccessible
a.cpp:15:40: error: within this context
make: *** [a] Error 1
\end{lstlisting}

When deriving a class in C++, the following situations can occur:
\begin{itemize}
\item Derive as \textit{public}: public and protected members of the base class remain public and protected members of the derived class.
\item Derive as \textit{protected}: public and protected members of the base class become protected members of the derived class.
\item Derive as \textit{private}: public and protected members of the base class become private members of the derived class.
\end{itemize}
In all cases, private members of the base class remain private, e.g.: not even the derived class can access them unless there is a specific friend declaration within the base class that explicitly grants access to them. If we do not provide an access specifier, the access rules of the baseclass are simply copied, e.g.: public remains public, protected remains protected and private remains private.

Unlike C++, Ruby does not support the changing of the access properties of members and methods of the superclass via one keyword (when we derive a class). However, Ruby has the strange property to allow the call to private methods and members from the superclass via the subclass. An example:

@@

This will print \textit{Random noise}. The explanation behind this is that Ruby actually allows to call private methods and members from a class via an implicit receiver. If you use an explicit receiver (for example \textit{Animal.speak}) the call will fail like you would expect. So although Ruby does not have an access specifier when deriving a class, it is however possible to mimic the behaviour from C++, even with more detail as you can control the access specifiers for all methods separatly. Only thing you have to do is rewrite the methods where you want some other access control and call the original method.

\subsection{Inheritance}
In Ruby, a class can only inherit from a single other class. Some other languages like C++ support multiple inheritance, a feature that allows classes to inherit features from multiple classes.
\begin{lstlisting}[language=Ruby]
class Animal
	def breath
		puts "Breathe"
	end
	def speak
		puts "Random noise"
	end
end

class Cat < Animal
	def speak
		puts "Meow"
	end
end
\end{lstlisting}
When using these two classes, one can see that methods will be overriden:
\begin{lstlisting}[language=Ruby]
cat = Cat.new
cat.speak
cat.breath
\end{lstlisting}
This will output:
\begin{lstlisting}
Random noise
Breathe
\end{lstlisting}

C++ supports multiple inheritance. With multiple inheritance it is possible to inherit from multiple classes like this:

@@

@@

When constructing an object of class D, C++ actually creates in the memory objects of B and C internally. B and C in their turn also construct each an object of A. This means that upon the creation of D, 4 internal objects are created: B, C and 2 times A. If we however have inherited B and C from A as virtual (for example: \textit{class B: virtual public A}), C++ will make sure that only one object of A is constructed.

@@

In this example, the class FooBar contains the content of module Foo (in this case the method Bar). It is possible to include multiple modules, even if the modules contain the same methods:

\begin{lstlisting}[language=Ruby]
module A
	def foo
		puts "Module A"
	end
end

module B
	def foo
		puts "Module B"
	end
end

class C
	include A
	include B
end

c = C.new
c.foo
\end{lstlisting}

This will output \textit{Module B}. The order in which a programmer includes modules, depends on what methods will be available. If we would have included A after B, the output would be \textit{Module A}. As modules are included, they override functionality that was defined in the lines before the include. Methods that are defined after the include statement, will override methods that were included.

\subsection{Polymorphism}
The literal meaning of polymorphism is "the ability to take on multiple forms or shapes". This refers, in a broader sense, to the ability of different objects to respond in different ways to the same message or method invocation. We have written an example in Ruby that shows what polymorphism can do:

\begin{lstlisting}[language=Ruby]
class Animal
	def makeNoise
		throw NotImplementedError.new("makeNoise() not implemented")
	end
end

class Dog < Animal
	def makeNoise
		puts "Woof!"
	end
end

class Cat < Animal
	def makeNoise
		puts "Meow!"
	end
end

[Dog, Cat, Animal].each do |obj|
	a = obj.new
	a.makeNoise
end
\end{lstlisting}

In this example, we see that the classes Dog and Cat are subclasses from Animal. We dynamically create a Dog and Cat object and call the method \textit{makeNoise}. The output is correctly:

@@

Even though Cat and Dog are different classes, they implement the same messages with a different implementation but can be called in exactly the same way. Via the implementation of \textit{makeNoise} in the class Animal, we make sure that there will be a clean error when someone has subclassed Animal but not overwritten \textit{makeNoise}; thus ensuring that all the subclasses of Animal understand the same message \textit{makeNoise} but with their own implementation. An equivalent example in C++ code:

\begin{lstlisting}
#include <iostream>

class Animal {
	public:
		virtual void makeNoise() = 0;
};

class Dog: public Animal {
	public:
		void makeNoise() {
			std::cout << "Woof!" << std::endl;
		}
};

class Cat: public Animal {
	public:
		void makeNoise() {
			std::cout << "Meow!" << std::endl;
		}
};

int main()
{
   Animal* a = new Dog();
   a->makeNoise();
   Animal* b = new Cat();
   b->makeNoise();
   return 0;
}
\end{lstlisting}

This correctly outputs:

@@

As you can see, we treat the objects Dog and Cat like they were Animal objects. We call the same method \textit{makeNoise()} which results in different results, e.g.: the same message results in a different execution.

\subsection{Namespaces/modules}
Namespaces in C++ are way to group classes, functions and other identifiers in one block. Outside this block, one must prefix the identifiers he wants to access with the namespace specifier. To ommit the prefixing, one can use the following statement:

@@

With this line, it is no longer necessary to prefix with \textit{abc::}. Namespaces are hierarchical, there can be mutiple namespaces defined in another namespace. When requesting an identifier, the compiler start searching from the namespace where the identifier is requested. For example if we request C in A::B, the compiler will first search if there exists a C in A::B. If not, the compiler searches in A and after that in the global namespace.

Modules are the Ruby equivalent of namespaces. They can also be defined hierarchical. The difference with C++ namespaces is that modules can be included in classes. This allows programmers to add new methods to classes. If several classes implement exact the same method, this method can be placed in a module to ensure that every class has the same code. This concept is called mixins which is an alternative in Ruby to multiple inheritance in C++.

@@

@@

Both C++ and Ruby do not support a keyword like \textit{interface}. It is however possible to enforce interfaces in these languages. In C++ this is done very easily: create an abstract class that only consists of virtual functions without a body. This enforces inherited classes to implement all the methods. If this is not done, the inherited classes will be abstract classes.

Via several modules, one can enforce interfaces in Ruby\footnote{An example of an implementation of interfaces can be found on http://www.metabates.com/2011/02/07/building-interfaces-and-abstract-classes-in-ruby/}. Generally, it is the same story as with abstract classes but you enforce the programmer to override all the methods of a class.

>>>>

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{color}
\usepackage{graphicx} 
\definecolor{codebg}{rgb}{0.9,0.9,0.9}
\lstset{language=C++,
backgroundcolor=\color{codebg},
numbers=left,
xleftmargin=9pt,
framexleftmargin=9pt,
basicstyle=\scriptsize,
numberstyle=\scriptsize\color{black},
breaklines=true,
numbersep=1pt,
tabsize=2}
\newenvironment{quoteje} 
               { \noindent {\bfseries  "}}     
               {{\bfseries  "}\vspace{\parskip}}
\renewcommand\refname{References}
\author{Jeroen Heymans}
\title{Comparison of two object-oriented languages: C++ and Ruby}
\begin{document}
\maketitle

@@

@@

\subsubsection{Ruby}
When designing Ruby, the creator of Ruby, Ykihiro Matsumoto, looked at other languages to find an ideal syntax. He wanted a dynamic typed scripting language that was more powerful than Perl and more object-oriented than Python. In Ruby, everything is an object. It was influenced by Smalltalk, every type was given method instance variables. All Ruby code was tested with Ruby version 1.9.2p0 (2010-08-18 revision 29036).
\pagebreak

\section{Comparisons}

@@

Due to the dynamic typed nature of Ruby, there is no need for constructions like templates in C++ to enforce one function or method to accept multiple types. Templates in C++ allow for creating generic functions, methods and classes. For example when we create a Stack class in Ruby, this can take any possible type of variable to put on the stack. In C++ however, because it is static typed, we need to notify upfront what the types of the elements on the stack will be. We could therefore write an \textit{IntegerStack}, \textit{FloatStack}, \textit{StringStack} and so on. With templates however, we can simply define a \textit{Stack\textless T\textgreater } where \textit{T} is the type that the stack will handle.

\subsection{Classes}
In object-oriented programming, a class is a blueprint that is used to construct objects which we call instances. We compare the several types of classes that can be made in Ruby and C++.

\subsubsection{Abstract classes}
An abstract class is a class that can not be instantiated for it is either labelled as abstract or it specifies abstract methods. The general idea of abstract classes is that the class must be extended. In C++, an abstract class is a class having at least one pure virtual function. An example can be found in listing 1 in the appendix.

If we are trying to compile C++ code where we instantiate an abstract class, we will receive an error message from the compiler that notifies us that this operation is not allowed:

@@

In order to use functionality defined in the abstract class, we need to inherit from the abstract class into another class where we override the virtual functions. If we do not override the virtual functions, the subclass will be an abstract class again.

Abstract classes in Ruby are a different story. Unlike some object-oriented languages, Ruby does not support a keyword like \textit{abstract} or a way like C++ to define an abstract class. Actually, Ruby does not provide the necessary standard tools to create abstract classes. However, the programmer himself can force a class to behave like an abstract class when he makes \textit{:new} (the constructor) private. The extending class must then make \textit{:new} public. This way, as long as there is no class that extends the class A with a private constructor, the methods in A will never be accesible as it is not possible to construct an instance of A, thus it behaves like an abstract class. The example code can be found in listing 2 in the appendix.

\subsubsection{Final classes}
A final class is a class that can not be extended. This could be done for reasons of security and efficiency. Final classes are not supported in C++ nor Ruby. However, there exist mechanisms to prevent programmers from extending other classes. An example of such a mechanism in C++ can be found in listing 3 of the appendix.

In this example, we made the class named Final final. We did this by creating a class Final that inherits from the class MakeFinal. In class MakeFinal, we have a private constructor which can only be called by the class Final as the class Final is marked as \textit{friend}. By inheriting MakeFinal \textit{virtual} in the class Final, we make sure that the constructor \textit{MakeFinal()} must be called if we derive the class Final. As only Final is a friend, this will generate a compile error that notifies us that \textit{MakeFinal()} is private if we would try to instantiate MakeFinal or to create an instance of MakeFinal. Via this mechanism, we have ensured that the class named Final is really final.

In Ruby, we can create a similar mechanism as can be found in listing 4 of the appendix. In this example, we add a method \textit{final} to the standardclass \textit{Class}. In this method \textit{final}, we define that a new method called \textit{inherited} should be added to the class given by the parameter \textit{klass}. If we then add a call to \textit{final} to our own defined class, the code for the method \textit{inherited} will be injected in our own class. \textit{inherited} is a method that will always be called whenever we try to inherit the class that contains this method. In our example, this will cause a \textit{RuntimeError}, thus disallowing the programmer to continue programming as long as we try to inherit from our \textit{final} class.

But is a class really final in Ruby? Not even if we force a class to be marked as final like we did in the example? Unfortunately not. This is due to the \textit{open classes} concept of Ruby that will always allow you to add methods to a class, whether it is marked as \textit{final} like in our example or not. See the section "Partial classes" however for an example on how to prevent extending an open class.

@@

\begin{itemize}
\item Friendships are not corresponded: if class A is a friend of class B, class B is not automatically a friend of class A.
\item Friendships are not transitive: if class A is a friend of class B and class B is a friend of class C, then is class A not a friend of class C.
\item Friendships are not inherited: if there is a class Base with a friend, then the class Derived (that inherits from Base) does not have that same friend. The same goes up for all friends of Derived, they are not automatically friends of Base. The same goes in the other direction: if Base (or Derived) is a friend of a class, then is Derived (or Base) not automatically a friend of the same class.
\item The restricted members and methods from Base that are inherited by Derived are accessible by a friend class of Derived if Derived itself can access these members and methods. So if Derived inherits publicly\footnote{See the section "Derived classes" in "Access control" for more details on deriving classes.} from Base, Derived only has access to the protected and public members from Base, not the private members, so neither does a friend (the same goes for methods).
\end{itemize}

Friend classes are not possible in standard Ruby, however there exist some libraries\footnote{An example can be found on https://github.com/lsegal/friend} that allow a programmer to define friend classes in Ruby.

\subsubsection{Inner classes}
Inner classes are called nested classes in Ruby. They can simply be defined like you would expect, by simply putting a class definition within a class definition as you can see in listing 6 in the appendix. If we have a class A with an inner class B, we can construct an object of the type \textit{A} for the outer class and \textit{A::B} for the inner class:

@@

With inner classes in C++, almost the same story exists. In order to access data from the inner class via the outer class or from the outer class via the inner class, an instance must be present to access this data. This can be achieved when constructing the inner or outer class as can be seen in listing 5 in the appendix.

@@

However, there are some rules concerning the access to and from the inner class in C++. The code from the listing will compile and work. But if you look closely, the method \textit{f()} from the class B can access member \textit{a} from class A eventhough \textit{a} is defined as private. So in fact, the inner class gets the privilege to act as if it is part of the outer class.

We must notice though that there is a different story when you want to access the inner class. In the code, A has a method \textit{f()} that can access the member \textit{test} that is defined in B, via the pointer to B that was saved in A. This is because the member \textit{test} is defined as \textit{public}. If we however define \textit{test} as \textit{protected} or \textit{private}, the compiler will notify us that this is not allowed.

@@

C++ does not have built-in support for metaclasses.

\subsubsection{Partial classes}
A partial class is a class whose definition may be splitted into multiple pieces. These pieces can be within one source file or across multiple files. Ruby supports the notion of partial classes. With this, it is possible to extend existing classes with new methods or members. Its not necessary that the class is user-defined, it can be for example the built-in \textit{String} class as you can see in listing 7 in the appendix. The result of executing \textit{doublePrint} on a String object, will give as a result: "\textit{testtest}".

Partial classes like in Ruby, do not exist in C++. It is not possible to add new methods or members on-the-fly. The reason for this is that Ruby classes are considered "open" while C++ classes are "closed". When you define a C++ class in the typical header-file, there is no possibility to split this definition in several files, e.g. multiple header files for the same class are not allowed. Only the definition of the methods themselves can be split over mutiple source-files. It is for example possible to have a header-file "a.h" and multiple source-files like "method1-from-a.cpp" and "method2-from-a.cpp". Still, this does not create pure open classes like in Ruby where you can add methods at runtime, it is merely a way to make smaller files.

If you do not want that classes are further extended in Ruby, then the possibility of \textit{freezing} the class exists. By simple adding the keyword \textit{freeze} to the class, you declare the class as frozen (e.g. not modifiable). From then on, the class is considered as \textit{closed} instead of the default \textit{open}. Classes in C++ could be regarded as standardly \textit{frozen}.

@@

Ruby also provides these three keywords but there are some subtle differences. We will now explain one difference, the rest will be explained in the following section. The difference between Ruby and C++ can be found in the objects on which a private method is called or a private member is requested. The C++ code that explains this can be found in listing 8 in the appendix and the Ruby code in listing 9 in the appendix. As you can see, we have a class A in which the \textit{test()}-method accepts a parameter to an A object. All three calls to the method \textit{foo()} work: with an implicit receiver and an explicit receiver like \textit{this} or \textit{a}. If we try the Ruby equivalent of this code, we get into trouble.

Both the lines \textit{self.foo} and \textit{obj.foo} generate the same error: \textit{NoMethodError}. This is because we use an explicit receiver. If you want to access a private member or method in Ruby, you can only use an implicit receiver. With any explicit receiver, even if that is \textit{self}, the \textit{NoMethodError} will be generated. We can conclude that Ruby depends on the type of the receiver if it will give you access to a private member or not.

In C++, this distinction does not exist. As long as you try to access the private member or method from code that is defined in the private member or method of his class, everything is okay. So if we have a private member \textit{test} in a class A, we can even access that member via an object instantiated from A that is not equal to \textit{this}, as long as the code is placed in the body of class A. Note: we have seen in the section "Friend classes" that friends of a class also have access to the private members and methods from a class.

\subsubsection{Derived classes}
When declaring a derived class in C++, we can provide an access specifier: \textit{public}, \textit{protected} or \textit{private}. For example if we have the base class A, we can create subclass B from A and provide one of the three access specifiers \textit{public}, \textit{protected} or \textit{private} like this:

\begin{lstlisting}
class A { };
class B: public A { };
class C: protected A { };
class D: private A { };
\end{lstlisting}

The purpose of this access specifier is to alter the access control of the members of the base class via the derived class. The access control is not altered on the original base class, only on the derived class. For example, if we derive a class as private, this does not mean that the methods and members of the original class become private, they keep their access specifier. If you want however to access a public member or method from a superclass via a subclass instance that was derived as private, you will receive an error. For an exmple, see listing 10 of the appendix.

When deriving a class in C++, the following situations can occur:
\begin{itemize}
\item Derive as \textit{public}: public and protected members of the base class remain public and protected members of the derived class.
\item Derive as \textit{protected}: public and protected members of the base class become protected members of the derived class.
\item Derive as \textit{private}: public and protected members of the base class become private members of the derived class.
\end{itemize}

In all cases, private members of the base class remain private, e.g.: not even the derived class can access them unless there is a specific friend declaration within the base class that explicitly grants access to them. If we do not provide an access specifier, the access rules of the baseclass are simply copied, e.g.: public remains public, protected remains protected.

Unlike C++, Ruby does not support the changing of the access properties of members and methods of the superclass via one keyword when we derive a class. This has to be done individually for every member or method.

We notice though, Ruby has the strange property to allow the call to private methods and members from the superclass via the subclass. An example:

@@

This will print \textit{Random noise}. The explanation behind this is that Ruby actually allows to call private methods and members from a class via an implicit receiver. If you use an explicit receiver (for example \textit{Animal.speak}) the call will fail like you would expect. So although Ruby does not have an access specifier when deriving a class, it is however possible to mimic the behaviour from C++, even with more detail as you can enable the access to private members and methods of the superclass. Ofcourse, overriding a method or member with a different access specifier is possible in C++. In case of a method, you have to call the superclass method with an explicit receiver. It is however not possible to gain access to private members or methods from the superclass, even by overriding the corresponding member or method.

\subsection{Inheritance}
In Ruby, a class can only inherit from a single other class. Some other languages like C++ support multiple inheritance, a feature that allows classes to inherit features from multiple classes.

We can define a superclass Animal and a subclass Cat in Ruby like we did in the listing number 11 in the appendix. When using these two classes, one can see that methods will be overriden:

\begin{lstlisting}[language=Ruby]
cat = Cat.new
cat.speak
cat.breath
\end{lstlisting}

This will output:

\begin{lstlisting}
Random noise
Breathe
\end{lstlisting}

This is classical single inheritance. C++ however supports multiple inheritance. With multiple inheritance it is possible to inherit from multiple classes like this:

@@

\begin{figure}
\caption{The diamond problem}
\begin{center}
\includegraphics[scale=0.6]{diamond.png}
\end{center}
\end{figure}

@@

When constructing an object of class D, C++ actually creates in the memory objects of B and C internally. B and C in their turn also construct each an object of A. This means that upon the creation of D, 4 internal objects are created: B, C and 2 times A. If we however have inherited B and C from A as virtual (for example: \textit{class B: virtual public A}), C++ will make sure that only one object of A is constructed. Since there is only one object of A, we now only have one possible method that was defined in A. If we want to access the methods from B or C, we simply have to prefix with respectively \textit{B::} or \textit{C::}.

@@

In this example, the class FooBar contains the content of module Foo (in this case the method Bar). It is possible to include multiple modules, even if the modules contain the same methods as you can see in listing x in the appendix.

Our code from listing x in the appendix will output "\textit{Module B}". The order in which a programmer includes modules, depends on what methods will be available. If we would have included A after B, the output would be \textit{Module A}. As modules are included, they override functionality that was defined in the lines before the include. Methods that are defined after the include statement, will override methods that were included.

\subsection{Polymorphism}
The literal meaning of polymorphism is "the ability to take on multiple forms or shapes". This refers, in a broader sense, to the ability of different objects to respond in different ways to the same message or method invocation. We have written an example in Ruby that shows what polymorphism can do, see listing 13 in the appendix. In our example, we see that the classes Dog and Cat are subclasses from Animal. We create a Dog and Cat object and call the method \textit{makeNoise}. The output is correctly:

@@

Even though Cat and Dog are different classes, they implement the same messages with a different implementation but can be called in exactly the same way. Via the implementation of \textit{makeNoise} in the class Animal, we make sure that there will be a clean error when someone has subclassed Animal but not overwritten \textit{makeNoise}; thus ensuring that all the subclasses of Animal understand the same message \textit{makeNoise} but with their own implementation. An equivalent example in C++ code can be found in listing 12 in the appendix. This code correctly outputs:

@@

As you can see, we treat the objects Dog and Cat like they were Animal objects. We call the same method \textit{makeNoise()} on an object that needs to be an Animal object. This results in different results, e.g.: the same message results in a different execution that is based on the actual type of the object.

Because Ruby is a duck typed language, it is even possible to disable the need for subclassing from the class Animal. It is perfectly possible to implement a class Car that is not subclassed from Animal and has a method called \textit{makeNoise} which results in printing "vroom". We can pass an instance of Car to the method print and it will work. The reason behind this, is that there is no real type checking in the background.

\begin{quote}
When I see a bird that walks like a duck and swims like a duck and quacks like a duck, I call that bird a duck.
\end{quote}

As long as the parameter of \textit{print} is an object that has a method \textit{makeNoise}, the interpreter of Ruby will not complain.

Such a system like in Ruby, can be simulated via the use of templates. With templates, you can for example make generic functions. If we take the example of a Cat and a Dog, we can modify the code so the \textit{print} function does not longer require that his only parameter is a pointer to an Animal-object. You can find the code in listing 14 in the appendix. By parameterizing this, we can enable the possibility to give an instance of the class Car to the print-function. This will work correctly as long as the given object has the methods that are called and the members that are requested or modified in the body of the templated function.

\subsection{Namespaces/modules}
Namespaces in C++ are a method to group classes, functions and other identifiers in one block. An example can be found in listing 15 in the appendix. Outside the block of a namespace, one must prefix the identifiers he wants to access with the namespace specifier. If you look at our example, if we want to access the integer \textit{test} that was defined within the namespace A, we need to prefix with \textit{A::} so we get: \textit{A::test}. To ommit the prefixing, one can use the following statement:

@@

With this line, it is no longer necessary to prefix with \textit{abc::}. Namespaces are hierarchical, there can be mutiple namespaces defined in another namespace. When requesting an identifier in the code in a namespace, the compiler start searching from the namespace where the identifier is requested. For example if we request C in code that is put in A::B, the compiler will first search if there exists a C in A::B. If not, the compiler searches in A and after that in the global namespace.

Modules are the Ruby equivalent of namespaces, an example can be found in listing 16 in the appendix. They can also be defined hierarchical. The difference with C++ namespaces is that modules can be included in classes. This allows programmers to add new methods to classes. If several classes implement exact the same method, this method can be placed in a module to ensure that every class has the same code. This concept is called mixins which is an alternative in Ruby to multiple inheritance in C++. It is discussed more in the section "Inheritance".

@@

When performing reflection, the rules defined via the access specifiers \textit{public}, \textit{protected} and \textit{private} are completely ignored. With reflection, it is therefore for example possible to get private members of an instance.

@@

Both C++ and Ruby do not support a keyword like \textit{interface}. It is however possible to enforce interfaces in these languages. In C++ this is done very easily: create an abstract class that only consists of virtual functions without a body, such a class is equivalent to an interface. This enforces inherited classes to implement all the methods. If an inherited class only implements a few of the virtual functions thus not all of them, the inherited class will be an abstract class. An example of this can be found in listing x in the appendix.

Via several modules, one can enforce interfaces in Ruby\footnote{An example of an implementation of interfaces can be found on http://www.metabates.com/2011/02/07/building-interfaces-and-abstract-classes-in-ruby/}. Generally, it is the same story as with abstract classes in Ruby but you enforce the programmer to override all the methods of a class.

\subsection{Method overloading}

C++ supports method overloading. It is possible to define multiple methods with the same name but with another arity or other types for the parameters. When compiling, the compiler will determine which method will be called during runtime.

Because Ruby is a dynamic typed language, method overloading is not possible.

\subsection{Constructors}

Constructors in C++ must have the name of the class as their name. If you want to define a constructor in the class \textit{Foo}, the constructor must be named \textit{Foo()}. Due to method overloading, it is possible to define more constructors where the types of the parameters or the arity may defer. Standardly, a compiler will make sure that there is a standard constructor in the class that accepts no arguments.
